---
title: "Alternate Color Projections"
author: "Will Murphy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using and Creating Other Color Projections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
library(knitr)
require(maps)
opts_chunk$set(fig.width = 6)
```

The YUV projection has been selected as the default for its interpretability,
but the colorplaner package framework can allow for other projections as well. 
This vignette explains the usage for the other color projections included in
the colorplaner package as well as instructions for creating custom color
projections. 

## The color_projection Option

The `color_projection` option in `scale_color_colorplane` and
`scale_fill_colorplane` can be used to alter how the mapped variables are 
mapped to color space. The argument can be provided either as a name of a
built-in color projection ("YUV", "red_blue", or "interpolate") or as a
function that will be used to perform the projection. Additional options to 
the color projection function can be provided to the scale function and will be 
passed through. For example, a value for `Y` can be provided in the scale call
to override the default for the fixed luminance level to be used in the YUV
projection. 

## Red/Blue Projection

The `red_blue_projection` option creates a scale wherein increasing values
for the horizontal variable are mapped on a white-to-red gradient and
increasing values for the vertical variable are mapped on a white-to-blue 
gradient. The colors from each gradient are blended by averaging the RGB
values so that the colors for large values of both variables are purple.

```{R redblue}
library(ggplot2)
library(colorplaner)

if(require(maps)) {
 crimes <- data.frame(state = tolower(rownames(USArrests)), USArrests)
  states_map <- map_data("state")
  ggplot(crimes,
         aes(map_id = state, fill = Murder, fill2 = UrbanPop)) +
    geom_map(map = states_map) +
    scale_fill_colorplane(color_projection = "red_blue") +
    expand_limits(x = states_map$long, y = states_map$lat) +
    coord_map() 
}
```

## Arbitrary Interpolation Projections

A generalized version of the average-of-gradients concept used in the red/blue
projection is available in `interpolate_projection`. By specifying the baseline
color (`zero_color`) and the colors to interpolate towards for each of the variables (`horizontal_color`, `vertical_color`), 
one can create customized color plane scales. The `interpolate_projection`
function will map each variable on a `colorRamp` from the baseline color to
the corresponding target color and blend the two gradient colors together by
averaging to determine the display color. 

```{R interpolate}
if(require(maps)) {
 crimes <- data.frame(state = tolower(rownames(USArrests)), USArrests)
  states_map <- map_data("state")
  ggplot(crimes,
         aes(map_id = state, fill = Murder, fill2 = UrbanPop)) +
    geom_map(map = states_map) +
    scale_fill_colorplane(color_projection = "interpolate",
                          zero_color = "darkorange2",
                          horizontal_color = "mediumspringgreen",
                          vertical_color = "#CD00CD") +
    expand_limits(x = states_map$long, y = states_map$lat) +
    coord_map()
}
```

## Creating New Color Projections

It is also possible to provide custom color projection functions to the 
colorplane scales. A custom projection function must accept as arguments two
numeric vectors and return a character vector of colors in `rgb` format 
(e.g., "#FFFFFF"). The function may safely assume that the input vectors will 
be of equal length, and the output vector must match this length. The input
vectors will represent the values of the horizontal and vertical mapped 
variables, but the will be scaled to the range of [0,1]. The projection
function does not need to handle missing values, these are handled at a higher
level in the scale and no missing values will be included in the vectors
passed to the function. The projection function can accept additional arguments
beyond the two numeric input vectors. In use, these should be provided as named
arguments in the `scale_color_colorplane`/`scale_fill_colorplane` call, and 
they will
be passed on to the color projection function when it is called. 

```{r, custom}
# Define a custom projection function that uses HSV color space by
# mapping the data to hue and saturation with a fixed level for value
hsv_projection <- function(x, y, v) {
  # Convert y value from a position to a hue angle
  h <- atan(scales::rescale(y, from = c(0,1), to = c(-1, 1)) / 
              scales::rescale(x, from =  c(0, 1), to = c(-1, 1)))
  # There are no missing values in the input, but atan can create some
  h <- ifelse(is.na(h), 0, h)
  h <- scales::rescale(h, from = c(-pi / 2, pi / 2), to = c(0, 1))
  # hsv takes inputs on a scale of [0, 1] and returns an RGB color string
  grDevices::hsv(h, x, v)
}

if(require(maps)) {
 crimes <- data.frame(state = tolower(rownames(USArrests)), USArrests)
  states_map <- map_data("state")
  ggplot(crimes,
         aes(map_id = state, fill = Murder, fill2 = UrbanPop)) +
    geom_map(map = states_map) +
    # The v argument will be passed on to hsv_projection
    scale_fill_colorplane(color_projection = hsv_projection, v = 0.75) +
    expand_limits(x = states_map$long, y = states_map$lat) +
    coord_map() 
}
````

The custom projection defined above creates a scale from HSV color space, yet
this scale is largely uninterpretable. Creating customized projections should 
be done with care to ensure the result is a meaningful and intuitive scale. 
